\documentclass[a4paper,11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\fontencoding{T1}
\usepackage{amsmath,amssymb,graphicx,color,enumerate,hyperref}
\title{Semantics of Causal Loop Diagrams\\ Draft}
\author{Sólrún Halla Einarsdóttir}
\date{\today}

\begin{document}
\maketitle
A Causal Loop Diagram (CLD) is a directed graph used to display causal
relationships between variables.
%
The vertices represent the variables and the edges represent qualitative
causal relationships, which can be positive or negative.

%We are interested in writing a DSL for CLDs in Haskell and propagating signs
%through CLDs using constraint programming, and would like to define a semantic that aids us in that.

We would like to define a semantic that aids us in reasoning about CLDs and will
describe two such semantics here.

\section{Notation}
We denote a positive causal relationship between $A$ and $B$ by
$A\xrightarrow{+} B$ and a negative one by $A \xrightarrow{-} B$.
%
Then $A \xrightarrow{+} B$ means that an increase in $A$ causes an
increase in $B$, while a decrease in $A$ causes a decrease in $B$.
%
On the other hand, $A\xrightarrow{-} B$ means that an increase in $A$
causes a decrease in $B$ (and conversely a decrease in $A$ causes an
increase in $B$).
%
We denote the sign of the edge from $A$ to $B$ by $s_{AB}$, so
$s_{AB}= +$ if $A\xrightarrow{+} B$ and $s_{AB}=-$ if
$A\xrightarrow{-} B$.

A vertex $A$ also has a sign $s_A$ that denotes the total influence on $A$, so
$s_A=+$ if there is an increase in $A$, $s_A=-$ if there is a decrease, $s_A=0$
if there is no change and $s_A=?$ if we cannot determine the change in $A$.

\section{QPN approach}

One approach to modelling and reasoning about CLDs is by using qualitative
probabilistic networks (QPNs).

A QPN is defined as a directed
acyclic graph $G=(V,E)$ where the vertices, $V$, correspond to
variables and the edges, $E$ to qualitative probabilistic influences.
%
These influences can be positive (+) or negative (-).
%
Later, when computing transitive influences we will also use (?) for
ambiguous and (0) for probabilistic independence

The meaning of signs on edges is defined according to first order
stochastic dominance, as follows:

Let $F_B(\cdot|a_i, x)$ be the cumulative distribution function (CDF) for
$B$ given $A=a_i$. Then $s_{AB}=+$ means that for all possible values
$a_1,a_2$ of $A$ where $a_1\geq a_2$, we must have:
\[F_B(b_0|a_1, x)\leq F_B(b_0|a_2, x),\]
that is,
\[P(B \leq b_0| A = a_1, x)\leq P(B\leq b_0| A = a_2, x)\]
for all possible values $b_0$ of $B$ and any consistent context $x$.
The context $x$ ranges over all possible assignments to the
variables other than $A$ that influence $B$, that are consistent with both
$A=a_1$ and $A=a_2$.
%
The definition of $s_{AB}=-$ is the same but with $a_1\leq a_2$.

In simpler terms, $s_{ab} = +$ means that greater values of $a$ mean
greater values of $b$ are more likely, and $s_{ab}=-$ means that
greater values of $a$ mean smaller values of $b$ are more likely.

These influences are symmetric, that is, $s_{xy}=s_{yx}$.
%
Due to this symmetry it is possible to propagate an observed increase
or decrease of one variable around the graph and find whether other
variables are then likely to have increased or decreased.

An overview of the subject can be found here:
\url{http://www.staff.science.uu.nl/~renoo101/Prof/Research/qpns.html}.
TODO: cite Wellman etc.

This definition is broad enough to
apply to many different systems and
be applicable to various real world situations.

We found some issues with QPNs that lead us to explore other
approaches.
%
First of all, since QPNs were originally defined for acyclic graphs
and the theory on them relies on acyclicity, they may not be the best
fit to describe CLDs, in which cycles (feedback loops) are an important
feature.
%
Inference on QPNs containing loops is difficult to implement and can
lead to ambiguous results.

Second, the formal semantics of inference on QPNs is difficult to
formalize since it relies heavily on not-so-simple probability theory.
%
(And stakeholder sessions will not produce data on probability
distributions.)

Additionally, as QPNs are defined solely based on qualitative
relationships there is no obvious way to expand them to also describe
quantitative relationships.

Lastly, since all inference in QPNs is probabilistic it leads to
results that may not be as meaningful or concrete as we would like,
such as ``there is a heightened probability that $x$ has increased'',
rather than ``$x$ has increased''.
%
For instance, a variable may decrease even though the cumulative
probabilistic influence on it is positive.

\section{Difference equation approach}

Inspired by a system of tanks with water flowing from one to another,
and in search of semantics that might also be extended to quantitative
reasoning, we came up with the following approach.

We consider the value nodes of the graph to be functions of the same
variable, such as a time variable $t$.

If we have a graph with two nodes, $x$ and $y$, and one edge from $x$
to $y$, then $s_{xy}=+$ implies that
\[\frac{\partial y}{\partial t} = F(x(t)),\]
where $F$ is a monotonically increasing function (monotonically decreasing for negative
causality, $s_{xy}=-$).

If the node $y$ has multiple parent nodes $x_1,\ldots,x_n$, then
$\frac{\partial y}{\partial t}$ depends on all the parent nodes. We can
isolate the effect of a single parent node $x$ on $y$ by differentiation.

In general we can then describe the causal relationship from $x$ to
$y$ as
\[\frac{\partial\left( \frac{\partial y(t)}{\partial t} \right)}{\partial x(t)} =
  f(x(t)),\]

where $f$ has a primitive function $F$ such that $F$ is monotonically
increasing if $s_{xy}=+$ and monotonically decreasing if $s_{xy}= -$.

This is somewhat more nuanced than $CLDs$ as they are described above,
where $s_{xy}=+$ implies that an increase in $x$ leads to an increase
in $y$, and a decrease in $x$ to a decrease in $y$.
%
Here we may have some threshold value $x_0$ for which $F(x_0) = 0$,
above which $x$ causes an increase in $y$, but an increase in $x$
causes a faster rate of increase in $y$ and a decrease in $x$ causes a
slowed rate of increase in $y$, and vice versa.

Note that though $F(x)$ is monotonically increasing, it may not be strictly
increasing, so we could for instance have $F(x) = 0$ for all $x < C$
for some threshold value $C$.

If the $y$ has parent nodes $x_1,\ldots,x_n$, then we have something
like
\[\frac{\partial y}{\partial t} = \sum_{i=1}F_i(x_i),\]
where $F_i$ is monotonically increasing if $s_{x_iy}=+$ but monotonically
decreasing if $s_{x_iy}=-$.

In a discrete time system we consider $\Delta(x_t) = x_t - x_{t-1}$
instead of $\frac{\partial x}{\partial t}$, and write
$\Delta(x_t) = F(y_{t-1})$ instead of
$\frac{\partial x}{\partial t} = F(y(t))$.
%
In simple cases we may only consider one time step with two values of
$t$, $t_{start}$ and $t_{end}$.

Here we explore how this approach relates to qualitative reasoning,
but it could be extended to quantitative reasoning by solving
appropriate differential equations.

\subsection{Simple qualitative model}

We consider a qualitative discrete time system where all values of
node variables are either +, -, 0, or ? (where ? is an ambiguous value assigned
to a variable whose value cannot be deduced). These values have the partial
ordering $- < 0 < +$, but ? cannot be compared to the other values. In place of
addition and multiplication we have the operations $\oplus$ and $\otimes$, whose
behavior can be seen in the following tables:
\begin{center}
\begin{tabular}{c|cccc}
$\oplus$ & + & - & 0 & ?\\
\hline
  +   & +  & ? & + & ?\\
  -   & ?  & - & - & ?\\
  0   & +  & - & 0 & ?\\
  ?   & ?  & ? & ? & ?\\
\end{tabular}
\quad
\begin{tabular}{c|cccc}
$\otimes$ & + & - & 0 & ?\\
\hline
  +   & +  & - & 0 & ?\\
  -   & -  & + & 0 & ?\\
  0   & 0  & 0 & 0 & 0\\
  ?   & ?  & ? & 0 & ?\\
\end{tabular}
\end{center}
The only
strictly increasing function in this system is the identity function $id(x) =
x$, and the only strictly decreasing function is the negation function $neg(x) =-\otimes x$.

For simplicity we consider the case where all initial values are
set to zero and $F_i(0)=0$, for all edges $i$. We only consider edge functions
$F_i$ where $F_i(?) = ?$, since we shouldn't be able to make unambiguous
deductions based on ambiguous values.
%

This is convenient for qualitative reasoning since then we are only
concerned with increases and decreases rather than numerical values.
%
The value of variable $x$ at time $t$, which we denote by $x_t$, then tells us whether there has been a net
increase or decrease in $x$.

Consider a graph with three nodes, a node $z$ and its two parent nodes
$x$ and $y$, $x\xrightarrow{s_{xz}} z$ and $y\xrightarrow{s_{yz}} z$.
%
Then we have
\[\Delta(z_t) = F_{xz}(x_{t-1}) \oplus F_{yz}(y_{t-1}),\]
where $F_{xz}$ and $F_{yz}$ are monotonically increasing or decreasing in
accordance with $s_{xz}$ and $s_{yz}$.
%
If we only allow strictly increasing/decreasing functions we then have
\[\Delta(z_t) = (s_{xz}\otimes x_{t-1})\oplus (s_{yz}\otimes y_{t-1}).\]

Consider a graph with three nodes $a$, $b$ and $c$ and two edges,
$a\xrightarrow{s_{ab}} b$ and $b\xrightarrow{s_{bc}} c$. Then we have
\begin{align*}
\Delta(c_t) &= F_{bc}(b_{t-1})\\
&= F_{bc}(b_{t-2} \oplus \Delta(b_{t-1}))\\
&= F_{bc}(b_{t-2}  \oplus F_{ab}(a_{t-2}))\\
\end{align*}

If $F_{bc}$ is linear, as is the case when we restrict the available functions
to the strictly increasing/decreasing $id$ and $neg$,
we then have
\[\Delta(c_t) = F_{bc}(b_{t-2})\oplus F_{bc}\circ F_{ab}(a_{t-2}).\]

If we only allow strictly increasing/decreasing functions we then have
\[\Delta(c_t) = (s_{bc}\otimes b_{t-2})\oplus (s_{bc}\otimes s_{ab}\otimes a_{t-2}),\]

\section{Relation to Generic Library}
We have modelled CLDs within the \verb|GenericLibrary|, the code for this model
can be seen in the file \verb|QualitativeExample.hs|. \\

% Say something further about formalizing semantics of GL using difference equations?
\end{document}
