The GRACeFUL project can be described as a stack of four layers,
starting with stakeholder interaction (group model building, WP2) at
the top, then the graphical user interface (GRACeFUL editor, WP3), the
domain specific language for models and constraints (GRACe DSL, WP4)
and finally the constraint programming (CP) backend providing
efficient solution algorithms (WP5).
%
In addition to the GRACe DSL, WP4 has developed a RESTful server around our DSL,
called GRACeServer, for handling webservice requests. The server communicates
both ``upwards'' to the GRACeFUL editor and ``downwards'' to the CP layer.
%
This deliverable describes our methods for verifying correctness of
this communication.

The division into frontend, server and backend is very common in
software systems and this means that our results should be applicable
for a wide range of other systems.
%
Our approach to software verification is based on three main parts:
%
declarative programming with strong types, property based testing in
general, and the SessionCheck tool for testing communicating systems
in particular.

\subsection{Scope and limitations}

This deliverable (D4.4) covers the software technology side of testing
and verification of RATs.
%
The CRUD RAT evaluation based on stakeholder sessions is reported
elsewhere [D2.6 Evaluation Report CRUD RATs: m36] as is the
interactivity and usability of the visual front-end [T3.4, D3.3 VA EDA
Tool Prototype (RAT components): m34].

\subsection{Rapid Assessment Tools}

Rapid assessment tools are used in large organisations like the World
Bank and the United Nations to assess risks and needs and to make
plans for improvement.
% http://www.wahooas.org/mshdvd2/assess_tools_MWL_Eng.htm
% http://www.fao.org/docrep/015/i2495e/i2495e06.pdf
In GRACeFUL the main focus is on Climate Resiliant Urban Design, but
the software developed in the project could be used for almost any
Global Systems Science problem.


\subsection{GRACeFUL software architecture}

\input{architecture}
